Reproducible Research: Peer Assessment 1
==========================================
Created by JORDI ROSES on October 19, 2014


### Basic settings

```{r}

echo = TRUE  # Always make code visible

options( scipen = 1 )  # Turn off scientific notations for numbers

```


### Loading and processing the data

* Load the data.
* Process/transform the data (if necessary) into a format suitable for your analysis.


```{r}

Current_Folder <- "C:/Users/jordi/Desktop/JORDI/COURSERA/REPRODUCIBLE RESEARCH/Peer Assessment 1/"

File_Name <- paste( Current_Folder , "activity.csv" , sep="" )

DATA <- read.csv( File_Name , colClasses = c( "integer" , "Date" , "factor" ) )




DATA$month <- as.numeric( format( DATA$date , "%m" ) )

DATA_noNA <- na.omit( DATA )

rownames( DATA_noNA ) <- 1 : nrow( DATA_noNA )

head( DATA_noNA )

dim( DATA_noNA )

```


### What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.


* Make a histogram of the total number of steps taken each day.

```{r}

library( ggplot2 )

ggplot( DATA_noNA , aes( date , steps ) ) + geom_bar( stat = "identity" , colour = "black" , fill = "black" , width = 0.5 ) + facet_grid(. ~ month , scales = "free" ) + labs( title = "Total Number of Steps Per Day" , x = "Date" , y = "Total Number of Steps" )

```


* Calculate and report the mean and median total number of steps taken per day.


MEAN / Total number of steps taken per day:

```{r}

totalSteps <- aggregate( DATA_noNA$steps , list( Date = DATA_noNA$date ) , FUN = "sum" )$x

mean( totalSteps )

```


MEDIAN / Total number of steps taken per day:

```{r}

median( totalSteps )

```


### What is the average daily activity pattern?

* Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

avgSteps <- aggregate( DATA_noNA$steps , list( interval = as.numeric( as.character( DATA_noNA$interval ) ) ) , FUN = "mean" )

names( avgSteps )[2] <- "meanOfSteps"

ggplot( avgSteps , aes( interval , meanOfSteps ) ) + geom_line( color = "black" , size = 0.5 ) + labs( title = "Time Series Plot of the 5-Minute Interval" , x = "5-Minute Intervals" , y = "Average Number of Steps Taken" )

```


* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

avgSteps[ avgSteps$meanOfSteps == max( avgSteps$meanOfSteps ) , ]

```


### Imputing missing values

Note that there are a number of days/intervals where there are missing values (NAs). The presence of missing days introduce bias into some calculations or summaries of the data.

* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs):


```{r}

sum( is.na( DATA ) )

```


* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.


We replace the NAs with the mean.


* Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

DATA_new <- DATA

for ( i in 1 : nrow( DATA_new ) ) {
    if ( is.na( DATA_new$steps[i] ) ) {
        
        DATA_new$steps[i] <- avgSteps[ which( DATA_new$interval[i] == avgSteps$interval ) , ]$meanOfSteps
        
    }
}

head( DATA_new )

sum( is.na( DATA_new ) )

```


* Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 


```{r}

ggplot( DATA_new , aes( date , steps ) ) + geom_bar( stat = "identity" ,
                                             colour = "black" ,
                                             fill = "black" ,
                                             width = 0.5 ) + facet_grid(. ~ month, scales = "free" ) + labs( title = "Histogram of Total Number of Steps Taken Each Day (no Missing Data)" , x = "Date" , y = "Total Number of Steps" )

```


* Do these values differ from the estimates from the first part of the assignment? 
* What is the impact of imputing missing data on the estimates of the total daily number of steps?


MEAN / Total number of steps taken per day:

```{r}

TotalSteps_new <- aggregate( DATA_new$steps , 
                           list( Date = DATA_new$date ) , 
                           FUN = "sum" )$x

Mean_new <- mean( TotalSteps_new )

Mean_new

```


MEDIAN / Total number of steps taken per day:

```{r}

Median_new <- median( TotalSteps_new )

Median_new

```


Comparative:

```{r}

Mean_old <- mean( totalSteps )
Median_old <- median( totalSteps )

Mean_new - Mean_old
Median_new - Median_old

```


There is no change in the means.
New median (with no NAs) is greater.



### Are there differences in activity patterns between weekdays and weekends?


* Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}

head( DATA_new )

DATA_new$weekdays <- factor( format( DATA_new$date , "%A" ) )

levels( DATA_new$weekdays )

levels( DATA_new$weekdays ) <- list( weekday = c( "Monday", "Tuesday" , "Wednesday" , "Thursday" , "Friday" ) , weekend = c( "Saturday" , "Sunday" ) )

levels( DATA_new$weekdays )

table( DATA_new$weekdays )

```


* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}

Steps_average <- aggregate( DATA_new$steps , 
                      list( interval = as.numeric( as.character( DATA_new$interval ) ) , 
                      weekdays = DATA_new$weekdays ) ,
                      FUN = "mean" )

names( Steps_average )[3] <- "MeanOfSteps"

library( lattice )

xyplot( Steps_average$MeanOfSteps ~ Steps_average$interval | Steps_average$weekdays , 
       layout = c( 1 , 2 ) , 
       type = "l" ,     
       xlab = "Interval" , 
       ylab = "Number of steps" )

```






